Preventing a Nuclear ISIL: \\Recommendations for United States Policy

### 2015-05-10 draft.tex
Added in Policy Section 6.7 - "Gift/Purchase".

\section{Post-Detonation Nuclear Forensics}

@@

\section{Purchase or Gift} \label{sec:purchase_gift}

\subsection{Export Control Regimes}

A significant concern identified earlier in this report is the possibility of a nuclear-capable state willingly selling a physics package to ISIL. The associated non-nuclear delivery technologies are also significant proliferation risks from states with the developed capabilities, if not more significant due to less oversight from international organizations and control regimes coupled with a higher degree of technology proliferation. Relevant delivery systems for WMD ISIL may consider include ICBMs, cruise missiles, and gravity bombs.

@@

\section{Weaponization to Target Selection}

If ISIL is able to obtain a nuclear weapon, the next steps for them would be to develop a method of delivery, pick a target, and deploy and detonate the bomb.  If possible, the first priority of the U.S. should be to locate where ISIL is storing the bomb and destroy it using drone strikes.  However, it is highly improbable that the exact location of the bomb will be known or found using current detection technologies.  Even if it is found, it might be stored underground where drone strikes would not reach it.  Therefore, the more practical strategy for the U.S. will be to try to stop ISIL at each of the previously mentioned steps.   First, the U.S. will need to determine what delivery methods are possible for ISIL.  The U.S. can prevent ISIL from acquiring the necessary components to build more complex delivery system and slow down movement of the bomb.  Then, the U.S. can make predictions about the group's targets based on geographical location from ISIL and try to intercept the bomb before it reaches the target.  If all of that fails and the weapon is deployed, all the U.S. can do is to help minimize the damage.  Then, the U.S. and IAEA must be prepared to collect nuclear forensic data after the detonation in order to acquire data to prevent the construction and deployment of a second bomb.

>>>>

\section{Post-Detonation Nuclear Forensics} \label{sec:post_det_forensics}

@@

\section{Gift/Purchase} \label{sec:purchase_gift}

Pathway: Gift/Purchase

Existing Policy: \textcolor{red}{Attribution through nuclear forensics}

Proposed Policy: \textcolor{red}{Declaratory policy regarding use of nuclear forensic attribution}

Lead Agency: DoD, DOS

Contributing Agency: DNDO, IC, NNSA, DTRA \normalfont

The targeted capability of concern for this policy is the ability of ISIL to obtain a complete nuclear weapon or components of a nuclear weapon via purchase or gift from a nuclear capable state.  Given our general lack of insight into foreign nations' nuclear weapons programs, locations, security, and accountability measures (at least in the open literature), it is impossible to develop a policy aimed at preventing the sale or gift of nuclear weapons or nuclear weapons components at the source.  Instead, the policy recommendation focuses on strengthening existing nuclear forensics capabilities to positively attribute any nuclear detonation to the country of origin while simultaneously issuing a declaratory policy on the U.S. response of the use of nuclear weapons against the U.S. or its allies.  Specifically, this declaratory policy would focus on the U.S. response to the transfer of nuclear technology, materials, or weapons that result in a detonation of a nuclear weapon on the U.S. or one of its allies.  This policy would close any potential loopholes in the Nuclear Posture Review's (NPR) declaratory policy and provide a tangible disincentive for the illicit proliferation of nuclear technology, materials, or weapons to ISIL \cite{U.S.DepartmentofDefense2010}.

Key to this policy option is the ability to definitely identify weapon characteristics including the country of origin, date of manufacture, and weapon design.  As discussed in \autoref{sec:post_det_forensics}, the U.S. and international capability maintains the ability to provide both pre- and post-detonation forensics of nuclear materials and weapon components that can be used in the attribution process.  The Nuclear Forensics and Attribution Act (NFAA) of 2010 provides guidance for the development and processing of nuclear forensics information, but it stops short of stating the end use of such information \cite{Act2010}.  The NFAA established the NTNFC within the DNDO \enquote{to provide centralized stewardship, planning, assessment, gap analysis, exercises, improvement, and integration for all federal nuclear forensics and attribution activities} \cite{Act2010}.  It also directs the DNDO to \enquote{establish a National Nuclear Forensics Expertise Development Program that is devoted to developing and maintaining a vibrant and enduring academic pathway from undergraduate to post-doctorate study in nuclear and geochemical science specialties directly relevant to technical nuclear forensics and that shall provide undergraduate and doctoral student scholarships and awards to ensure that faculty and their graduate students have a sustained funding stream} \cite{Act2010}.  Without explicitly naming organizations, the NFAA also calls for the President to \enquote{develop protocols for the data exchange and dissemination of sensitive information relating to nuclear or radiological materials and samples} \cite{Act2010}.  Some of the organizations that will provide relevant information and samples for nuclear forensics are the NNSA, DTRA, and the IC.

The proposed policy  would strengthen the commitment to nuclear forensics by the NFAA through increased funding for the program from current \$20M 2015 budget to \$30M for 2016 \cite{Year2015}.  This money would be used for nuclear forensics-related research, to \enquote{provide steady funding streams} for nuclear engineers and physicists, to develop forensics capabilities in the area of post-detonation debris collection, and increase the number, fidelity, and participation in nuclear forensic exercises.  In addition to the increased funding, the proposed policy  would have the President, in coordination with the DOS and DoD, issue a declaratory policy regarding the U.S. response to the transfer of nuclear technology, materials, or weapons that result in a detonation of a nuclear weapon on the U.S. or one of its allies.  The proposed declaratory policy would be incorporated into the next NPR, but issued as a standalone document in the interim due to the dynamic ISIL threat.  The proposed policy wording is:

\blockquote{The United States is prepared to strengthen its strategic and extended deterrence guarantees by declaring that any state, regardless of NPT obligations compliance, that facilitates or allows the transfer of nuclear material, knowledge, or weapons to sub-national groups will face devastating United States military response up to and including the use of nuclear weapons.  Any individuals responsible for the illicit transfer, whether national leaders or military commanders, would be held fully accountable.}

@@

\section{Weaponization and Delivery}

If ISIL is able to obtain a nuclear weapon, the next steps for them would be to develop a method of delivery, pick a target, and deploy and detonate the bomb.  If possible, the first priority of the U.S. should be to locate where ISIL is storing the bomb and destroy it using drone strikes.  However, it is highly improbable that the exact location of the bomb will be known or found using current detection technologies.  Even if it is found, it might be stored underground where drone strikes would not reach it.  Therefore, the more practical strategy for the U.S. will be to try to stop ISIL at each of the previously mentioned steps.   First, the U.S. will need to determine what delivery methods are possible for ISIL.

The associated non-nuclear delivery technologies are also significant proliferation risks from states with the developed capabilities, if not more significant due to less oversight from international organizations and control regimes coupled with a higher degree of technology proliferation. Relevant delivery systems for WMD ISIL may consider include ICBMs, cruise missiles, and gravity bombs.

The U.S. can prevent ISIL from acquiring the necessary components to build more complex delivery system and slow down movement of the bomb.  Then, the U.S. can make predictions about the group's targets based on geographical location from ISIL and try to intercept the bomb before it reaches the target.
